---
title: "T-cell input RDS objects, and matching ProjecTILs predictions back to cells"
author: "alzheimers pipeline"
date: "2026-06-03"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Heavy chunks that read the multi-GB donor RDS are eval=FALSE by default.
# Run them deliberately in an interactive session on a box with headroom.
```

> **Memory warning — read this first.**
> The two donor objects are **~5 GB each on disk** and expand well past that
> in RAM. This is a shared box. Do **not** `readRDS()` either one "just to
> look." Every heavy chunk below is marked `eval=FALSE`; run them one at a
> time, immediately `DietSeurat()` to RNA-only, and `rm(); gc()` before moving
> on. The *prediction* CSVs, by contrast, are a few MB — load those freely.

## 0. What we're walking through

| Role | File | Size |
|---|---|---|
| Donor1 query | `data/datasets/tcells/donor1/scrna/Tcells.singlet.rds` | 5.2 GB |
| Donor2 query | `data/datasets/tcells/donor2/scrna/Tcells_d2.singlet (1).rds` | 4.6 GB |
| CD8 reference atlas | `data/external/projectils/CD8T_human_ref_v1.rds` | 246 MB |
| CD4 reference atlas | `data/external/projectils/CD4T_human_ref_v1.rds` | 250 MB |
| Per-cell labels (output) | `data/derived/tcells_incytr_inputs/<donor>/scrna/projectils_predictions.csv` | 1–1.4 MB |

The donor RDS are the *only* place individual cells live. The pipeline never
mutates them — it reads each once, projects onto the CD8/CD4 atlases with
ProjecTILs, and writes one row per cell to `projectils_predictions.csv`. The
**barcode** (`colnames(obj)`) is the single key that ties that CSV back to a
cell. The whole point of §4 is to show that join is exact and 1:1.

Authoritative scripts (this vignette mirrors them, it does not replace them):

- `alz/ingest/tcells_projectils_map.R` — produces the predictions.
- `alz/ingest/tcells_scrna_extract.R` — consumes them (the canonical join).
- `alz/incytr_pair/build_tcells_seurat.R` — stamps the sanitized label as `Type`.

```{r libs, eval=FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Matrix)
})
# Run from the repo root so the relative paths below resolve.
# setwd("~/Projects/work/alzheimers")
```

---

## 1. Per-donor config (the only thing that differs)

The two donors came from different runs, so one metadata column name differs.
Everything else in this vignette is donor-agnostic.

```{r cfg, eval=FALSE}
cfg <- list(
  donor1 = list(
    rds     = "data/datasets/tcells/donor1/scrna/Tcells.singlet.rds",
    day_col = "sample_ID"),     # donor1 stores the timepoint here
  donor2 = list(
    rds     = "data/datasets/tcells/donor2/scrna/Tcells_d2.singlet (1).rds",
    day_col = "Sample_Label")   # donor2 stores it here
)
donor <- "donor1"               # flip to "donor2" to redo the walkthrough
this  <- cfg[[donor]]
```

Note the donor2 filename keeps the collaborator's literal `" (1)"` — the
scripts reference it verbatim; do not rename it.

---

## 2. Load one object, memory-safely

The pipeline pattern: load once, drop to the RNA assay, kill `scale.data`,
then never re-read. We do the same.

```{r load, eval=FALSE}
t0  <- Sys.time()
obj <- readRDS(this$rds)
cat("read in", round(as.numeric(Sys.time() - t0, units = "secs"), 1), "s\n")
cat("dim (genes x cells):", paste(dim(obj), collapse = " x "), "\n")

DefaultAssay(obj) <- "RNA"
obj <- DietSeurat(obj, assays = "RNA", dimreducs = NULL, graphs = NULL)
obj[["RNA"]]$scale.data <- NULL   # the single biggest dense block
gc(full = TRUE)
```

---

## 3. Inspect the structure

Don't assume the metadata schema — discover it. These are all cheap once the
object is in memory.

```{r structure, eval=FALSE}
# 3a. Class, assays, default assay
class(obj)
Assays(obj)
DefaultAssay(obj)

# 3b. Cell and feature counts
ncol(obj)                      # cells  (donor1 ~25,678 ; donor2 ~20,654)
nrow(obj)                      # genes

# 3c. Barcode format — THIS is the join key. 10x-style, with a "-1" suffix.
head(colnames(obj))
#> "AAACCAAAGCCAGATC-1" "AAACCAAAGCCCTTGC-1" ...

# 3d. Full metadata schema (column names + types)
str(obj@meta.data, max.level = 1)

# 3e. The two columns the pipeline actually reads:
#     the day column (name differs per donor) and seurat_clusters
table(obj@meta.data[[this$day_col]])     # raw timepoint labels
table(obj$seurat_clusters)               # Louvain/Leiden clusters (NOT the label)

# 3f. Expression layers present (need a log-normalized 'data' layer downstream)
SeuratObject::Layers(obj[["RNA"]])
```

**Key facts to internalize here:**

- **`seurat_clusters` is *not* the cell-type label.** It's the unsupervised
  graph clustering that shipped in the object. The biological state comes only
  from ProjecTILs (§4–5). The cluster-keyed path was deleted 2026-05-28.
- **The day column name differs** (`sample_ID` vs `Sample_Label`); both encode
  the timepoint as a string like `...Day17...`, parsed with
  `sub(".*[Dd]ay[_ ]?(\\d+).*", "\\1", x)`.
- **The barcode carries no donor/day prefix** — it's a bare 10x barcode. Days
  are recovered from the metadata column, not the barcode.

### The reference atlases (optional, ~250 MB each — safe to peek)

```{r refs, eval=FALSE}
cd8 <- readRDS("data/external/projectils/CD8T_human_ref_v1.rds")
sort(unique(as.character(cd8$functional.cluster)))
#> "CD8.CM" "CD8.EM" "CD8.MAIT" "CD8.NaiveLike" "CD8.TEMRA" "CD8.TEX" "CD8.TPEX"
rm(cd8); gc()
```

These define the *vocabulary* of `functional.cluster` — the raw state strings
that land in the prediction CSV (before sanitization in §5).

---

## 4. Matching `projectils_predictions.csv` back to individual cells

This is the heart of the question. **The join key is `barcode == colnames(obj)`,
and the CSV has exactly one row per cell of the original object** — including
cells ProjecTILs could not call (their `functional.cluster` is `NA`). So the
match is total and 1:1; nothing is pre-filtered out of the CSV.

### 4a. Load the (small) prediction table

```{r pred-load}
donor    <- "donor1"
predpath <- file.path("data/derived/tcells_incytr_inputs",
                      donor, "scrna", "projectils_predictions.csv")
pred <- read.csv(predpath, stringsAsFactors = FALSE, check.names = FALSE)

str(pred)
head(pred)
#> barcode            seurat_clusters day lineage_gate functional.cluster  ...conf
#> AAACCAAAGCCAGATC-1 6               17  CD8          CD8.TPEX            0.536
#> AAACCAAAGGAATGGT-1 12              0   none         NA                  NA
```

Columns:

| column | meaning |
|---|---|
| `barcode` | the cell barcode — **the key** (= `colnames(obj)`) |
| `seurat_clusters` | copied from the object, for reference only |
| `day` | timepoint parsed from the donor's day column |
| `lineage_gate` | scGate outcome: `CD8`, `CD4`, `doublet`, or `none` |
| `functional.cluster` | raw ProjecTILs state; `NA` when `lineage_gate ∈ {none, doublet}` |
| `functional.cluster.conf` | projection confidence; `NA` alongside an `NA` state |

### 4b. Prove the key is 1:1 with the object (run with `obj` loaded)

```{r pred-verify, eval=FALSE}
obj_bc  <- colnames(obj)
pred_bc <- pred$barcode

# (i) same set of barcodes, no extras on either side
setequal(obj_bc, pred_bc)          # expect TRUE
length(obj_bc) == nrow(pred)       # expect TRUE  (one CSV row per cell)

# (ii) no duplicate barcodes in the CSV
!any(duplicated(pred_bc))          # expect TRUE

# (iii) CSV order need NOT match object order — so ALWAYS join by name,
#       never by position. Build a named lookup keyed on barcode:
state_by_bc <- setNames(pred$functional.cluster,      pred$barcode)
gate_by_bc  <- setNames(pred$lineage_gate,            pred$barcode)
conf_by_bc  <- setNames(pred$functional.cluster.conf, pred$barcode)
```

### 4c. Attach the predictions onto the cells

`AddMetaData` aligns by the rownames of the supplied frame, so re-index the
prediction table to the object's barcode order first. This is the safe,
canonical way to put the label on each cell.

```{r pred-attach, eval=FALSE}
rownames(pred) <- pred$barcode
pred_aligned   <- pred[colnames(obj), ]      # reorder rows to match cells exactly
stopifnot(identical(rownames(pred_aligned), colnames(obj)))

obj <- AddMetaData(obj, metadata = pred_aligned[, c(
  "lineage_gate", "functional.cluster", "functional.cluster.conf")])

# Now every cell carries its ProjecTILs call:
head(obj@meta.data[, c("seurat_clusters", "lineage_gate", "functional.cluster")])
table(obj$lineage_gate, useNA = "ifany")
```

The equivalent one-liner the pipeline actually uses (no reordering needed
because it indexes the named vector directly — see
`tcells_scrna_extract.R:98-101`):

```{r pred-attach-pipeline, eval=FALSE}
obj$functional.cluster <- setNames(pred$functional.cluster, pred$barcode)[colnames(obj)]
obj$projectils_gate    <- setNames(pred$lineage_gate,       pred$barcode)[colnames(obj)]
```

### 4d. Inspect specific cells end-to-end

```{r pred-spotcheck, eval=FALSE}
# Pull the raw counts + the assigned label for one barcode of interest
bc <- "AAACCAAAGCCAGATC-1"
obj@meta.data[bc, c("seurat_clusters", this$day_col,
                    "lineage_gate", "functional.cluster", "functional.cluster.conf")]

# All cells called CD8.TEX (exhausted CD8) at day 20:
sel <- which(obj$functional.cluster == "CD8.TEX" & obj$ts_day == 20)
length(sel)
colnames(obj)[head(sel)]
```

---

## 5. From raw ProjecTILs state to the label used downstream

The CSV stores the **raw** atlas state (`CD8.TEX`, `CD4.CTL_Exh`, …). Incytr
splits identifiers on `_`, so the pipeline sanitizes each raw state to an
alphanumeric label via a fixed 14-entry map. This is the label that becomes the
cell's `Type` in `build_tcells_seurat.R` and the state shown in the viewer.

```{r label-map}
LABEL_MAP <- c(
  "CD8.CM" = "CD8CM", "CD8.EM" = "CD8EM", "CD8.MAIT" = "CD8MAIT",
  "CD8.NaiveLike" = "CD8Naive", "CD8.TEMRA" = "CD8TEMRA",
  "CD8.TEX" = "CD8Tex", "CD8.TPEX" = "CD8Tpex",
  "CD4.CTL_EOMES" = "CD4CTLeomes", "CD4.CTL_Exh" = "CD4CTLexh",
  "CD4.CTL_GNLY" = "CD4CTLgnly", "CD4.NaiveLike" = "CD4Naive",
  "CD4.Tfh" = "CD4Tfh", "CD4.Th17" = "CD4Th17", "CD4.Treg" = "Treg"
)

# Apply to the prediction table; cells with NA state stay NA (they get dropped)
raw   <- pred$functional.cluster
mask  <- !is.na(raw)
# Guard: an atlas state not in the map is a hard error upstream, not a silent pass
stopifnot(length(setdiff(unique(raw[mask]), names(LABEL_MAP))) == 0)
pred$state <- NA_character_
pred$state[mask] <- LABEL_MAP[raw[mask]]

table(pred$state, useNA = "ifany")
```

**Cells dropped here are dropped honestly:** any cell with `lineage_gate ∈
{none, doublet}` has `functional.cluster = NA`, gets no `state`, and is removed
before aggregation (donor1 ~13%, donor2 ~7%). The drop accounting is written to
`state_audit.json` (`n_total`, `n_kept`, `drop_by_gate`).

```{r audit}
audit <- jsonlite::fromJSON(file.path(
  "data/derived/tcells_incytr_inputs", donor, "scrna", "state_audit.json"))
audit[c("n_total", "n_kept", "n_dropped", "drop_pct")]
audit$drop_by_gate
```

---

## 6. Cross-checks you can run without the big RDS

Even without loading a donor object, the small derived files let you confirm the
labels are internally consistent:

```{r crosschecks}
donor <- "donor1"
d <- function(f) file.path("data/derived/tcells_incytr_inputs", donor, "scrna", f)

pred  <- read.csv(d("projectils_predictions.csv"), check.names = FALSE)
cc    <- read.csv(d("cell_counts.csv"))             # (state, day, n_cells)
audit <- jsonlite::fromJSON(d("state_audit.json"))

# (a) labeled-cell count from the raw predictions == audit n_kept
sum(!is.na(pred$functional.cluster) &
    !(pred$lineage_gate %in% c("none", "doublet")))   # vs audit$n_kept

# (b) cell_counts totals reconcile to the audit per-state totals
aggregate(n_cells ~ state, cc, sum)

# (c) embeddings (if present) are a subset of barcodes that passed a gate
emb <- tryCatch(read.csv(d("projectils_embeddings.csv"), check.names = FALSE),
                error = function(e) NULL)
if (!is.null(emb)) all(emb$barcode %in% pred$barcode)   # expect TRUE
```

---

## 7. Summary — the mental model

```
 Tcells.singlet.rds                 projectils_predictions.csv
 ┌───────────────────┐  ProjecTILs  ┌──────────────────────────────┐
 │ cell = colname     │ ───────────▶ │ barcode  (== colname)        │  one row
 │ counts/data (RNA)  │  CD8 + CD4   │ lineage_gate  {CD8/CD4/...}  │  per cell,
 │ meta: sample_ID,   │  scGate +    │ functional.cluster (raw)     │  incl. the
 │       seurat_clust │  projection  │ functional.cluster.conf      │  uncalled
 └───────────────────┘              └──────────────────────────────┘  (NA) ones
        ▲                                          │
        │  AddMetaData / setNames(...)[colnames]   │  LABEL_MAP (14 states)
        └───────── join on `barcode` ─────────────▶│  → sanitized `state` → Type
```

- **One key, one direction:** `barcode == colnames(obj)`, 1:1, order-independent
  — always join by name.
- The CSV is the complete per-cell record (uncalled cells included as `NA`); the
  *drop* happens downstream, recorded in `state_audit.json`.
- `seurat_clusters` ≠ the cell-type label; the biological state is ProjecTILs'
  `functional.cluster`, sanitized through `LABEL_MAP`.
- Re-run order if you ever regenerate: `tcells-projectils-map` →
  `tcells-scrna-extract` → `tcells-build-incytr-seurat`.
